---
title: "Exercise 1 C: Visium HD Segmented Data"
format: html
editor: source
editor_options:
  chunk_output_type: console
execute:
  warning: false
---

## Learning Objectives

By the end of this exercise, you will be able to:

-  TO DEFINE
<!-- Adapt depending on the final data structure -->


## Libraries

```{r day1-1c-visium_HD_segmented-1}
#| message: false
#| warning: false
#| output: false

library(sf)
library(arrow)
library(dplyr)
library(tidyr)
#library(scuttle)
#library(SingleR)
#library(Voyager)
library(ggplot2)
#library(ggspavis)
#library(VisiumIO)
#library(OSTA.data)
#library(patchwork)
#library(SpotSweeper)
#library(DropletUtils)
library(SpatialExperiment)
library(SpatialFeatureExperiment)
```

## Data for the course

<!-- Adapt depending on the final data structure -->

```{r day1-1c-visium_HD_segmented-3}
options(timeout = 600)

if (!dir.exists("data/Visium_HD_Human_Colon_Cancer_segmented_outputs/")) {
  dir.create("data", showWarnings = FALSE)
  download.file(
    url = "https://cf.10xgenomics.com/samples/spatial-exp/4.0.1/Visium_HD_Human_Colon_Cancer/Visium_HD_Human_Colon_Cancer_segmented_outputs.tar.gz",
    destfile = "data/Visium_HD_Human_Colon_Cancer_segmented_outputs.tar.gz",
    mode = "wb"
  )

  untar(
    tarfile = "data/Visium_HD_Human_Colon_Cancer_segmented_outputs.tar.gz",
    exdir = "data/"
  )

  file.remove("data/Visium_HD_Human_Colon_Cancer_segmented_outputs.tar.gz")
} else {
  message("Data exists, please proceed to next steps!")
}
```

We start reading the object into a `SpatialExperiment` object and then subset it to the course region, as we have done for visium HD binned data in the previous exercises.


```{r day1-1c-visium_HD_segmented-4}
spe_seg <- TENxVisiumHD(
  segmented_outputs = "data/segmented_outputs/",
  processing = "filtered",
  format = "h5",
  images = "lowres"
) |>
  import()

rownames(spe_seg) <- uniquifyFeatureNames(
  ID = rowData(spe_seg)$ID,
  names = rowData(spe_seg)$Symbol
)

spe_seg
```



##########

Once the object is loaded, define the region of interest (ROI) for the course. 


```{r day1-1c-visium_HD_segmented-5}
roi_visium <- c(
  xmin = 50000,
  xmax = 58000,
  ymin = 14500,
  ymax = 22000
)
roi_visium

spe_seg <- spe_seg[, spatialCoords(spe_seg)[, 1] >= roi_visium[["xmin"]] &
  spatialCoords(spe_seg)[, 1] <= roi_visium[["xmax"]] &
  spatialCoords(spe_seg)[, 2] >= roi_visium[["ymin"]] &
  spatialCoords(spe_seg)[, 2] <= roi_visium[["ymax"]]]
```



Load the previous binned object for comparison:

```{r day1-1c-visium_HD_segmented-6}
spe <- loadHDF5SummarizedExperiment(dir="results/day1/", prefix="01.1_spe")
```

::: callout-important
## Exercise 1

- 1. Compare the number of features and spots in the segmented and binned objects. What do you observe?
- 2. Compare the colData of the two objects. What are the differences in the metadata columns? What do they represent?
- 3. Do you observe any other differences in the content of the objects? For example, in the rowData or the spatial coordinates?

:::

<!-- Adapt the questions depending on the final order of exercises, add answers -->

```{r day1-1c-visium_HD_segmented-7}
spe_seg
plotVisium(spe, zoom=T)
plotVisium(spe_seg, zoom=T)

plotVisium(spe,
  annotate = "PIGR",
  assay = "counts",
  zoom = TRUE,
  point_size = 1,
  point_shape = 22)

plotVisium(spe_seg,
  annotate = "PIGR",
  assay = "counts",
  zoom = TRUE,
  point_size = 1,
  point_shape = 22)


```

Using visium HD segmented data allows us to have a more detailed view of the tissue, as we are not limited to the coordinates of the spots. The x and y coordinates of the `SpatialExperiment` object will correspond to the centroid of the segmented cells, but we can also have access to the full segmentation of the cells.

Information about the cell segmentation is stored in the `metadata` of the `SpatialExperiment` object, and it contains the coordinates of the boundaries of each segmented cell. In order to visualize the segmentation, we creare a `SpatialFeatureExperiment` object, which allows us to store and visualize spatial features such as the cell segmentation boundaries (see `colGeometries` function in the `SpatialFeatureExperiment` package).

<!-- Adapt depending on the final order of exercises -->

```{r day1-1c-visium_HD_segmented-8}
sfe_seg <- toSpatialFeatureExperiment(spe_seg)
colGeometries(sfe_seg)

# Convert numeric cell_id (e.g. 3) -> "cellid_000000003-1"
cell_id_to_colname <- function(x) {
  paste0("cellid_", sprintf("%09d", as.integer(x)), "-1")
}
cell_id_char <- cell_id_to_colname(metadata(spe_seg)$cellseg$cell_id)

# add cell segmentation boundaries
seg <- metadata(spe_seg)$cellseg
rownames(seg) <- cell_id_char

i <- colnames(spe_seg)
j <- match(i, rownames(seg))
seg <- seg[j, ]
all(colnames(sfe_seg) == rownames(seg))
colGeometries(sfe_seg) <- list(cellseg=seg)
colGeometries(sfe_seg)
```

We can visualise such geometries using 
```{r day1-1c-visium_HD_segmented-9}

plot(st_geometry(colGeometry(sfe_seg)), col=rep(colors(), 2), )

```  

<!--- JUNA: y-axis is flipped using plot() --->

Save the object
```{r day1-1c-visium_HD_segmented-9}
saveHDF5SummarizedExperiment(sfe_seg,
  dir = "results/day1", prefix = "01.1_sfe_seg", replace = TRUE,
  chunkdim = NULL, level = NULL, as.sparse = NA,
  verbose = NA
)

```  

